Concurrent Programming with CXL

### 2020-06-17 book/paper.tex
Latest

\begin{itemize}
\item In Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} introduces the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter;
\item Chapter~\ref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a critical
section, and Chapter~\ref{ch:naive} presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant implementation of a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} introduces the notion of \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the bounded buffer
problem and various solutions.
\item Chapter~\ref{ch:starvation} talks about the problem that in some
synchronization solutions, processes may not be able to get access to a
resource they need, and a solution applied to the reader/writer problem.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
condition variables, another approach to process synchronication.
\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
avoid them.
\item Chapter~\ref{ch:actor} presents the actor model as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} shows solutions to process synchronization
that prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} introduces barrier synchronization, useful in
high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

\section*{Exercises}
\begin{enumerate}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual is violated, the assertion should fail, and a trace should be reported.
Now plug \texttt{while True: pass;;} before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.  Now mutual exclusion is guaranteed but
progress is violated.  CXL should print a trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports.
\item If we change the loop in Figure~\ref{fig:cs} into an infinite loop,
does the \emph{progress} property guarantee that all processes that are trying
to enter eventually enter?  Why or why not?
\end{enumerate}

\chapter{Na\"{\i}ve Attempts at Implementing Critical Sections}
\label{ch:naive}

@@

\begin{quote}
Try your own solutions!
\begin{itemize}
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{itemize}
\end{quote}

@@

This anecdote suggests the following.  If you need to do an induction
proof of an algorithm, you have to come up with an inductive invariant.
Before trying to prove the algorithm, you can check that the predicate is
at least invariant by testing it using CXL.  Doing so could potentially
avoid wasting your time on a proof that will not work because the
predicate is not invariant, and therefore not an inductive invariant either.

To finish the story, the author of CXL fixed the Wikipedia page, but there
is another cool story.  Years later a colleague of the author, teaching
the same course, asked if the first two assignments (setting \texttt{flags[self]})
to \texttt{True} and \texttt{turn} to \texttt{1 - self}) can be reversed.
After all, they are different variables assigned independent values---in a
sequential program one could surely swap the two assignments.

\begin{quote}
What do you think?
\begin{itemize}
\item See if you can figure out for yourself if the two assignments can be
reversed.  Then run the program in Figure~\ref{fig:peterson} after reversing
the two assignments and see what happens.
\end{itemize}
\end{quote}

\begin{quote}
Bonus question!
\begin{itemize}
\item Can you generalize Peterson's algorithm to more than two processes?
\end{itemize}
\end{quote}

@@

\begin{quote}
\begin{itemize}
\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
per process rather than a simple counter?
\item Can you write a version of this test that uses a counter instead
of a flag per process?
\item Write a ``library implementation'' of reader/writer locks so multiple
reader/writer locks can be instantiated.
\end{itemize}
\end{quote}

>>>>

\begin{itemize}
\item In Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} introduces the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter;
\item Chapter~\ref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a critical
section presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} introduces the notion of \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the bounded buffer
problem and various solutions.
\item Chapter~\ref{ch:starvation} talks about the problem that in some
synchronization solutions, processes may not be able to get access to a
resource they need, and a solution applied to the reader/writer problem.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
condition variables, another approach to process synchronication.
\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
avoid them.
\item Chapter~\ref{ch:actor} presents the actor model as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} shows solutions to process synchronization
that prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} introduces barrier synchronization, useful in
high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

@@

\section*{Exercises}
\begin{enumerate}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual is violated, the assertion should fail, and a trace should be reported.
Now plug \texttt{while True: pass;;} before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.  Now mutual exclusion is guaranteed but
progress is violated.  CXL should print a trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports.
\item If we change the loop in Figure~\ref{fig:cs} into an infinite loop,
does the \emph{progress} property guarantee that all processes that are trying
to enter eventually enter?  Why or why not?
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{enumerate}

@@

This anecdote suggests the following.  If you need to do an induction
proof of an algorithm, you have to come up with an inductive invariant.
Before trying to prove the algorithm, you can check that the predicate is
at least invariant by testing it using CXL.  Doing so could potentially
avoid wasting your time on a proof that will not work because the
predicate is not invariant, and therefore not an inductive invariant either.
To finish the story, the author of CXL fixed the Wikipedia page.

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/csonebit.cxl}
\end{code}
\caption{[csonebit.cxl] Mutual exclusion using a flag per process.}
\label{fig:csonebit}
\end{figure}

\section*{Exercises}
\begin{enumerate}
\item Figure~\ref{fig:csonebit} presents another correct solution to the
mutual exclusion problem.  It is similar to the one in
Figure~\ref{fig:upflags}, but has a process \emph{back out and try again}
if it finds that the other process is either trying to enter the critical
section or already has.  Compare this algorithm with Peterson's.  What are
the advantages and disadvantages?
\item A colleague of the author asked if the first two assignments in
Peterson's algorithm (setting \texttt{flags[self]})
to \texttt{True} and \texttt{turn} to \texttt{1 - self}) can be reversed.
After all, they are different variables assigned independent values---in a
sequential program one could surely swap the two assignments.
See if you can figure out for yourself if the two assignments can be
reversed.  Then run the program in Figure~\ref{fig:peterson} after reversing
the two assignments and describe in English what happens.
\item Bonus question:
Can you generalize Peterson's algorithm to more than two processes?
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
per process rather than a simple counter?
\item Can you write a version of this test that uses a counter instead
of a flag per process?
\item Write a ``library implementation'' of reader/writer locks so multiple
reader/writer locks can be instantiated.
\end{enumerate}
